Efficient Constructions of Prefer-Max Sequences and of Infinite de Bruijn Sequences

### 2017-01-06 paper.tex
...

% Theorem like env.
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{reptheorem}{Theorem}

@@

The observation made in the last theorem directed us to examine the reverse of the prefer-max sequence. Specifically, the structure identified in Theorem~\ref{thm:onion} means that if we manage to construct the sequence backwards, we do not have to specify $k$ ahead of time. We can add symbols forever while $k$ grows in layers: we first write $0^n$, then a word whose substrings of length $n$ are $\{0,1\}^n$, then extend with other words in $\{0,1,2\}^n$, and so on.

The second contribution of this work is then an efficient construction of this reversed sequence. This is formulated in Algorithm~\ref{inf-construction}. We say that this construction is efficient because each step can be computed in $O(n^2)$ operations by enumerating all the $n$ rotations of $\sigma w$ and by comparing them lexicographically (reading each word from right to left).

@@

\setcounterref{reptheorem}{thm:onion}
\addtocounter{reptheorem}{-1}
\begin{reptheorem}[Onion]
	The last $(k-1)^n+1$ symbols of the $(k,n)$-prefer-max sequence form the $(k-1,n)$-prefer-max sequence.
\end{reptheorem}
\begin{proof}
	Denote the symbols of the $(n-k$)-prefer-max sequence by $s_1,s_2,\dots,s_{k^n+1}$ and let, for notational convenience, $s_i=0$ for all $i<1$. Let ${i_0} = \min\bigl\{i\colon (k-1) \notin \{s_{i-n+1},\dots,s_{i}\}\bigr\}$. 
	By minimality we have that $s_{{i_0}-n}=k-1$. Since $s_{i_0} \neq k-1$ and because $s_{{i_0}-n+1}\cdots s_{{i_0}-1} (k-2)$ did not appear as a substring before ${i_0}$ (it does not contain the symbol $k-1$) we get, by construction, that $s_{{i_0}}=k-2$. This means that there exists some ${i_1}<{i_0}$ such that $s_{{i_1}-n+1}\cdots s_{{i_1}} = s_{{i_0}-n+1}\cdots s_{{i_0}-1}(k-1)$.

	Assume, towards contradiction, that $s_{{i_0}-n+1}\cdots s_{{i_0}-1} \neq 0^{n-1}$. In particular $s_{{i_1}-n+1}\cdots s_{{i_1}} \neq 0^{n-1}(k-1)$ which means that ${i_1} > 1$. Therefore, $k-1  \in \{s_{{i_1}-n},\dots,s_{{i_1}-1}\}=\{s_{{i_1}-n}\} \cup \{s_{{i_0}-n+1},\dots,s_{{i_0}-1}\}$. Since the second set in this union does not contain $k-1$, we get that $s_{{i_1}-n}=k-1$. This leads to a contradiction because it means that $s_{{i_1}-n}\cdots s_{{i_1}-1} = s_{{i_0}-n}\cdots s_{{i_0}-1}$ which cannot happen in a de Bruijn sequence. This contradiction gives us that $s_{{i_0}-n+1}\cdots s_{{i_0}-1} = 0^{n-1}$.

	The sequence is therefore as follows:
	$$\underbrace{s_1,\dots,s_{i_0-n-1},k-1,{0,\dots,0}}_{\parbox{10 em}{\scriptsize \centering{all substrings of length $n$ contain $k-1$}}},\underbrace{k-2}_{s_{i_0}},s_{i_0+1},\dots,s_{k^n+1}$$
	To complete the proof, we need to show that all substrings of length $n$ that contain $k-1$ appear before $s_{i_0}$.

	For $w \in \{0,\dots,k-1\}^n$ and $1 \leq m \leq n$, let $j_m$ be such that $s_{j_m-n+1}\cdots s_{j_m} = w[m..n]0^{m-1}$, where $w[m..n]$ is the $n-m+1$ letters suffix of $w$. 
	Because $0$ is the smallest symbol in our alphabet, we get that it is added only if all the other symbols cannot be added, i.e., when adding other symbols would generate a substring that has already being seen before. 
	Since $w$ comes, as a window, right before some $w[2..m]\tau$,
	we have that $w[2..m]0$ comes after (or is equal to) $w[2..m]\tau$ which is after $w$. Similarly we can see that $w[2..m]0$ appears before $w[3..m]0^2$ and so on.
	This gives us that $j_1 < j_2 < \cdots < j_n$. Note that $s_{j_1-n+1}\cdots s_{j_1}=w$, i.e, that $w$ appears as a substring at the $j_1$'s symbol of the sequence.

	We will now show that if $w$ contains the symbol $k-1$ it must appear as a substring before the $i_0$'s symbol in the sequence. Consider first the case where the last (right-most) symbol of $w$ is $k-1$. In this case, $j_n=i_{0}-1$ so we have, because $j_1 < j_n$, that $w$ appears as a substring before the $i_0$'th symbol in the sequence. The second case is when the last (right-most) symbol in $w$ is not $k-1$ but there is another letter in $w$ which is $k-1$. In this case we have that $w$ appears less than $n$ steps after a window that ends with $k-1$ and the window $s_{i_0-n}\cdots s_{i_0-1}=(k-1)0^{n-1}$ does not appear in the windows between them because they all contain $k-1$ at a letter that is not  their first. Thus, also in this case, the window $w$ appears before the index $i_0$.

	To conclude the proof we note that we established that the windows up to the index $i_0$ contain all and only the words in $\{0,\dots,k-1\}^n \setminus \{0,\dots,k-1\}^n$ and that this part of the sequence ends with $n-1$ zeros and then the next part begins with the symbol $k-2$. This means that from this point on the prefer-max construction acts exactly as it does when the sequence begins with $k-2$ so it must constructs the $(k-1,n)$-prefer-max sequence.
\end{proof}

>>>>

% Theorem like env.
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{lemma}[theorem]{Lemma}

\newtheorem{reptheorem}{Theorem}
\newenvironment{repeatedtheorem}[1]
{
	\setcounterref{reptheorem}{#1}
	\addtocounter{reptheorem}{-1}
	\begin{reptheorem}
	}{			
	\end{reptheorem}
}

@@

The observation made in the last theorem directed us to examine the reverse of the prefer-max sequence. Specifically, the structure identified in Theorem~\ref{thm:onion} means that, if we manage to construct the sequence backwards, we do not have to specify $k$ ahead of time. Instead, we can add symbols forever while $k$ grows in layers: we first write $0^n$, then a word whose substrings of length $n$ are $\{0,1\}^n$, then extend with other words in $\{0,1,2\}^n$, and so on.

The second contribution of this work is then an efficient construction of this reversed sequence. This is formulated in Algorithm~\ref{inf-construction}. We say that this construction is efficient because each step can be computed in $O(n^2)$ operations by enumerating all the $n$ rotations of $\sigma w$ and comparing them lexicographically (reading each word from right to left).

@@

\begin{repeatedtheorem}{thm:onion}[Onion]
	{The last $(k-1)^n+1$ symbols of the $(k,n)$-prefer-max sequence form the $(k-1,n)$-prefer-max sequence.}
\end{repeatedtheorem}
\begin{proof}
	Denote the symbols of the $(n-k$)-prefer-max sequence by
	$s_1,s_2,\dots,s_{k^n+1}$ and let, for notational convenience, $s_i=0$ for all
	$i<1$.  For each $i\geq 1$ let $w_i=s_{i-n+1}\cdots s_i$ be the ``sliding window''
	observed at the $i$th symbol of the sequence. 
	The proof goes by showing that $\{w_1,\dots,w_{k^n-(k-1)^n}\} = \{0,\dots,k-1\}^n \setminus \{0,\dots,k-2\}^n$.

	Let ${i_0} = \min\{i\colon k-1 \notin w_i\}$. 
	By minimality, because  $s_{{i_0}-n}$ is the only symbol that is in $w_{i_0-1}$ and is not in $w_{i_0}$, 
	we have that $s_{{i_0}-n}=k-1$. Since $s_{i_0} \neq k-1$ and
	because $s_{{i_0}-n+1}\cdots s_{{i_0}-1} (k-2) \notin \{w_1,\dots,w_{i_0-1}\}$ because it does not contain the symbol $k-1$, 
	we get, by the construction of the prefer-max sequence,
	that $s_{{i_0}}=k-2$. This means that there exists some ${i_1}<{i_0}$ such that
	$w_{i_1} = s_{{i_0}-n+1}\cdots s_{{i_0}-1}(k-1)$.

	Assume, towards contradiction, that $s_{{i_0}-n+1}\cdots s_{{i_0}-1} \neq
	0^{n-1}$. In particular $w_{i_1} \neq 0^{n-1}(k-1)$ which
	means that ${i_1} > 1$. Therefore, $k-1  \in w_{i_1-1}=
	\{s_{{i_1}-n},\dots,s_{{i_1}-1}\}=\{s_{{i_1}-n}\} \cup
	\{s_{{i_0}-n+1},\dots,s_{{i_0}-1}\}$. Since the second set in this union does
	not contain $k-1$, we get that $s_{{i_1}-n}=k-1$. This leads to a contradiction
	because it means that $w_{{i_1}-1} = w_{{i_0}-1}$ which cannot happen in a de Bruijn sequence. This contradiction
	gives us that $s_{{i_0}-n+1}\cdots s_{{i_0}-1} = 0^{n-1}$.

	The sequence is therefore as follows:
	$$\underbrace{s_1,\dots,s_{i_0-n-1},k-1,\overbrace{0,\dots,0}^\text{$n\text{-}1$ times}}_{\text{all the $w_i$s here contain
				$k-1$}},\underbrace{k-2}_{s_{i_0}},s_{i_0+1},\dots,s_{k^n+1}$$
	To complete the argument, we need to show that all substrings of length $n$ that
	contain $k-1$ appear before $s_{i_0}$.

	For $w \in \{0,\dots,k-1\}^n$ and $1 \leq m \leq n$, let $j_m$ be such that
	$w_{j_m} = w[m..n]0^{m-1}$, where $w[m..n]$ is the $n-m+1$
	letters suffix of $w$. 
	Because $0$ is the smallest symbol in our alphabet, we get that it is added
	only if all the other symbols cannot be added, i.e., when adding other symbols
	would generate a substring that has already being seen before. 
	Since $w$ comes, as a window, right before some $w[2..m]\tau$,
	we have that $w[2..m]0$ comes after (or is equal to) $w[2..m]\tau$ which is
	after $w$. Similar arguments give us that $w[2..m]0$ appears before $w[3..m]0^2$ and
	so on. Together we get that $j_1 < j_2 < \cdots < j_n$. Note that $w_{j_1}=w$, 
	i.e, that $w$ appears as a substring at the $j_1$'s symbol of the
	sequence.

	We will now show that if $w$ contains the symbol $k-1$ it must appear as a
	substring before the $i_0$'s symbol in the sequence. Consider first the case
	where the last (right-most) symbol of $w$ is $k-1$. In this case, $j_n=i_{0}-1$
	so we have, because $j_1 < j_n$, that $w$ appears as a substring before the
	$i_0$'th symbol in the sequence. The second case is when the last (right-most)
	symbol in $w$ is not $k-1$ but there is another letter in $w$ which is $k-1$. In
	this case we have that $w$ appears less than $n$ steps after a window that ends
	with $k-1$ and the window $w_{i_0-1}=(k-1)0^{n-1}$ does not
	appear in the windows between them because they all contain $k-1$ at a letter
	that is not  their first. 
	Formally, we have that there is some $m<n$ such that 
	$j_1-m < i_0$,  $w \in \{w_{{j_1}-m},\dots,w_{j_1-1}\}$ and 
	$w_{i_0-1} \notin \{w_{{j_1}-m},\dots,w_{j_1-1}\}$. This gives us that, also in this 
	case, the window $w$ appears before	the index $i_0$.

	To conclude the proof we note that we established that the windows up to the
	index $i_0$ contain all and only the words in $\{0,\dots,k-1\}^n \setminus
	\{0,\dots,k-2\}^n$ and that this part of the sequence ends with $n-1$ zeros and
	then the next part begins with the symbol $k-2$. This means that from this point
	on the prefer-max construction acts exactly as it does when the sequence begins
	with $k-2$ so it must constructs the $(k-1,n)$-prefer-max sequence.
\end{proof}
